---
title: "Reproducible Example of Self-Updating Process Clustering"
output: 
  rmarkdown::html_vignette:
    number_sections: yes
    toc: yes
bibliography: supc.bib
author: Wush Wu
vignette: >
  %\VignetteIndexEntry{supc}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

This document shows how to reproduce the simulations and examples in @doi:10.1080/00949655.2015.1049605.

# Applications

## Seeds Data

## Golub Data

The data presented in @Golub99molecularclassification shows the strength of the self-updating process that can isolate noises.

First, we load the pre-processed and normalized dataset from package `multtest`(@multtest):

```{r load-golub, eval = FALSE}
data(golub, package = "multtest")
```

### Discover gene patterns

In @doi:10.1080/00949655.2015.1049605, the parameter is: `r rawToChar(as.raw(c(0x60, 0x72, 0x20, 0x3d, 0x20, 0x34, 0x2e, 0x36, 0x60)))`. So we execute the function `supc1` with parameter `tau = 4.6`:

```{r golub-discover1, eval = FALSE}
```

